\begin{frame}
	\frametitle{Diversity with Unison}

	Implemented 3 different strategies

	\vspace{0.5cm}

	\textit{Cost}, frequency of surviving gadgets, execution time of code generator

\end{frame}

\begin{frame}
	\frametitle{Strategies}

	\begin{enumerate}
		\item Enumerate
			\begin{itemize}
				\item Do nothing when a solution is found
				\item The same combination is never explored twice
			\end{itemize}
		\item Registers
			\begin{itemize}
				\item Disallow the same register allocation
			\end{itemize}
		\item Schedule
			\begin{itemize}
				\item Disallow same schedule
				\item Remember there can be more or fewer instructions
			\end{itemize}
	\end{enumerate}

\end{frame}

\begin{frame}
	\frametitle{Data set}

	Unison test suite

	\vspace{0.5cm}

	23 functions; All samples from SPEC2006

	\vspace{0.5cm}

	Simulate linking by placing them in the same order every time


\end{frame}

\begin{frame}
	\frametitle{Experimental Protocol}

	\begin{enumerate}
		\item For every function generate 1000 versions with "speed" as the optimization goal.
		\item Version 0 of every function will make up program 0, version 1 will make up program
			1 and so forth.
		\item For every program find all gadgets and its cost.
		\item Calculate our metrics.
		\item Repeat for all strategies.
			\begin{itemize}
				\item Enumerate.
				\item Registers.
				\item Schedule.
			\end{itemize}
		\item Repeat for all sampling rates.
			\begin{itemize}
				\item 1.
				\item 10.
				\item 100.
				\item 1000.
			\end{itemize}
	\end{enumerate}

\end{frame}
